---
title: "Regression vs. PCA"
subtitle: "*DSAN 5300: Statistical Learning*"
date: last-modified
institute: "<a href=\"mailto:jj1088@georgetown.edu\">`jj1088@georgetown.edu`</a>"
author: "Jeff Jacobs"
categories: ["Extra Writeups"]
bibliography: "../../_DSAN5300.bib"
weeknum: 2
assignment: "General"
cache: true
format:
  revealjs:
    output-file: slides.html
    footer: "DSAN 5300 Extra Slides: Regression vs. PCA"
    scrollable: true
    df-print: kable
    code-fold: true
    slide-number: true
    html-math-method: mathjax
    css: "../../dsan-globals/jjstyles.css"
    include-in-header:
      text: "<link href=\"https://fonts.googleapis.com/css2?family=Material+Symbols+Outlined\" rel=\"stylesheet\" />"
    revealjs-plugins:
      - simplemenu
    simplemenu:
      flat: true
      barhtml:
        header: >
          <div class='menubar'>
            <span style='position: absolute; left: 8; padding-left: 8px;'><a href='./index.html'>&larr; Return to Notes</a></span>
            <ul class='menu'></ul>
          </div>
      scale: 0.5
  html:
    output-file: index.html
    df-print: kable
    code-fold: true
    warning: false
    include-in-header: {"text": "<link href=\"https://fonts.googleapis.com/css2?family=Material+Symbols+Outlined\" rel=\"stylesheet\" />"}
---

## The Central Tool of Data Science {data-name="What is Regression?"}

::: {.hidden}

```{r}
#| label: r-source-globals
source("../../dsan-globals/_globals.r")
```

:::

{{< include ../../dsan-globals/_globals-tex.qmd >}}

* If science is understanding relationships between variables, **regression** is the most basic but fundamental tool we have to start **measuring** these relationships
* Often exactly what humans do when we see data!

::: columns
::: {.column width="48%"}
```{r}
#| label: basic-scatter
library(ggplot2)
library(tibble)
x_data <- seq(from=0, to=1, by=0.02)
num_x <- length(x_data)
y_data <- x_data + runif(num_x, 0, 0.2)
reg_df <- tibble(x=x_data, y=y_data)
ggplot(reg_df, aes(x=x, y=y)) +
  geom_point(size=g_pointsize) +
  dsan_theme("quarter")
```
:::

::: {.column width="4%"}
<div style="height: 50%; display: flex; flex-direction: column; align-items: center; justify-content: center;">
<p>
<span class="material-symbols-outlined" style="visibility: hidden;">
psychology
</span>
<span class="material-symbols-outlined gsym-medium" style="transform: translate(0px, 20px);">
psychology
</span><br>
<span class="material-symbols-outlined gsym-medium">
trending_flat
</span>
</p>
</div>
:::

::: {.column width="48%"}

```{r}
#| label: basic-regression
ggplot(reg_df, aes(x=x, y=y)) + 
  geom_point(size=g_pointsize) +
  geom_smooth(method = "lm", se = FALSE, color = cbPalette[1], formula = y ~ x, linewidth = g_linewidth*3) +
  dsan_theme("quarter")
```
:::
:::

## The Goal

* Whenever you carry out a regression, keep the **goal** in the front of your mind:

::: {.callout-tip .r-fit-text title="The Goal of Regression"}

**Find a line $\widehat{y} = mx + b$ that best predicts $Y$ for given values of $X$**

:::

## How Do We Define "Best"? {.smaller}

* Intuitively, two different ways to measure **how well a line fits the data**:

::: {layout="[1,1]" layout-valign="center"}

```{r}
#| label: pc-line
#| fig-width: 6
N <- 11
x <- seq(from = 0, to = 1, by = 1 / (N - 1))
y <- x + rnorm(N, 0, 0.25)
mean_y <- mean(y)
spread <- y - mean_y
df <- tibble(x = x, y = y, spread = spread)
ggplot(df, aes(x=x, y=y)) +
  geom_abline(slope=1, intercept=0, linetype="dashed", color=cbPalette[1], linewidth=g_linewidth*2) +
  geom_segment(xend=(x+y)/2, yend=(x+y)/2, linewidth=g_linewidth*2, color=cbPalette[2]) +
  geom_point(size=g_pointsize) +
  coord_equal() +
  dsan_theme("half") +
  labs(
    title = "Principal Component Line"
  )
```

```{r}
#| label: reg-line
#| fig-width: 6
ggplot(df, aes(x=x, y=y)) +
  geom_abline(slope=1, intercept=0, linetype="dashed", color=cbPalette[1], linewidth=g_linewidth*2) +
  geom_segment(xend=x, yend=x, linewidth=g_linewidth*2, color=cbPalette[2]) +
  geom_point(size=g_pointsize) +
  coord_equal() +
  dsan_theme("half") +
  labs(
    title = "Regression Line"
  )
```

:::

::: {.aside}
On the difference between these two lines, and why it matters, I cannot recommend @gelman_data_2007 enough!
:::

## Principal Component Analysis {.smaller .crunch-title .crunch-ul .shift-footnotes-20}

* **Principal Component Line** can be used to **project** the data onto its **dimension of highest variance**

* More simply: PCA can **discover** meaningful axes in data (**unsupervised** learning / **exploratory** data analysis settings)

```{r}
#| label: gdp-plot
#| warning: false
#| fig-width: 10
#| fig-height: 4
library(readr)
library(ggplot2)
gdp_df <- read_csv("assets/gdp_pca.csv")

dist_to_line <- function(x0, y0, a, c) {
    numer <- abs(a * x0 - y0 + c)
    denom <- sqrt(a * a + 1)
    return(numer / denom)
}
# Finding PCA line for industrial vs. exports
x <- gdp_df$industrial
y <- gdp_df$exports
lossFn <- function(lineParams, x0, y0) {
    a <- lineParams[1]
    c <- lineParams[2]
    return(sum(dist_to_line(x0, y0, a, c)))
}
o <- optim(c(0, 0), lossFn, x0 = x, y0 = y)
ggplot(gdp_df, aes(x = industrial, y = exports)) +
    geom_point(size=g_pointsize/2) +
    geom_abline(aes(slope = o$par[1], intercept = o$par[2], color="pca"), linewidth=g_linewidth, show.legend = TRUE) +
    geom_smooth(aes(color="lm"), method = "lm", se = FALSE, linewidth=g_linewidth, key_glyph = "blank") +
    scale_color_manual(element_blank(), values=c("pca"=cbPalette[2],"lm"=cbPalette[1]), labels=c("Regression","PCA")) +
    dsan_theme("half") +
    remove_legend_title() +
    labs(
      title = "PCA Line vs. Regression Line",
	    x = "Industrial Production (% of GDP)",
	    y = "Exports (% of GDP)"
    )
```

::: {.aside}
See <a href="https://juliasilge.com/blog/un-voting/" target="_blank">https://juliasilge.com/blog/un-voting/</a> for an amazing blog post using PCA, with 2 dimensions, to explore UN voting patterns!
:::

## Create Your Own Dimension!

```{r}
#| label: pca-plot
#| warning: false
ggplot(gdp_df, aes(pc1, .fittedPC2)) +
    geom_point(size = g_pointsize/2) +
    geom_hline(aes(yintercept=0, color='PCA Line'), linetype='solid', size=g_linesize) +
    geom_rug(sides = "b", linewidth=g_linewidth/1.2, length = unit(0.1, "npc"), color=cbPalette[3]) +
    expand_limits(y=-1.6) +
    scale_color_manual(element_blank(), values=c("PCA Line"=cbPalette[2])) +
    dsan_theme("full") +
    remove_legend_title() +
    labs(
      title = "Exports vs. Industrial Production in Principal Component Space",
      x = "First Principal Component (Dimension of Greatest Variance)",
      y = "Second Principal Component"
    )
```

## And Use It for EDA

```{r}
#| label: pca-facet-plot
#| warning: false
library(dplyr)
library(tidyr)
plot_df <- gdp_df %>% select(c(country_code, pc1, agriculture, military))
long_df <- plot_df %>% pivot_longer(!c(country_code, pc1), names_to = "var", values_to = "val")
long_df <- long_df |> mutate(
  var = case_match(
    var,
    "agriculture" ~ "Agricultural Production",
    "military" ~ "Military Spending"
  )
)
ggplot(long_df, aes(x = pc1, y = val, facet = var)) +
    geom_point() +
    facet_wrap(vars(var), scales = "free") +
	dsan_theme("full") +
	labs(
		x = "Industrial-Export Dimension",
		y = "% of GDP"
	)
```

## But in Our Case... {.crunch-title .crunch-ul data-name="Regression Hypothesis"}

* $x$ and $y$ dimensions **already have meaning**, and we have a **hypothesis** about $x \rightarrow y$!

::: {.callout-tip title="The Regression Hypothesis $\mathcal{H}_{\text{reg}}$"}
Given data $(X, Y)$, we estimate $\widehat{y} = \widehat{\beta_0} + \widehat{\beta_1}x$, hypothesizing that:

* Starting from $y = \widehat{\beta_0}$ when $x = 0$ (the intercept),
* An increase of $x$ by 1 unit is associated with an increase of $y$ by $\widehat{\beta_1}$ units (the coefficient)
:::

* We want to measure **how well** our line predicts $y$ for any given $x$ value $\implies$ **vertical distance** from regression line

## Key Features of Regression Line {.crunch-math .smaller-math}

* Regression line is **BLUE**: **B**est **L**inear **U**nbiased **E**stimator
* What exactly is it the "best" linear estimator of?

$$
\widehat{y} = \underbrace{\widehat{\beta_0}}_{\small\begin{array}{c}\text{Predicted} \\[-5mm] \text{intercept}\end{array}} + \underbrace{\widehat{\beta_1}}_{\small\begin{array}{c}\text{Predicted} \\[-4mm] \text{slope}\end{array}}\cdot x
$$

is chosen so that

::: {.smallmath}

$$
\theta = \left(\widehat{\beta_0}, \widehat{\beta_1}\right) = \argmin_{\beta_0, \beta_1}\left[ \sum_{x_i \in X} \left(\overbrace{\widehat{y}(x_i)}^{\small\text{Predicted }y} - \overbrace{\expect{Y \mid X = x_i}}^{\small \text{Avg. }y\text{ when }x = x_i}\right)^2 \right]
$$

:::

## Regression in `R` {.crunch-output}

```{r}
#| label: lin-reg
#| echo: true
lin_model <- lm(military ~ industrial, data=gdp_df)
summary(lin_model)
```

## `lm` Syntax

```r
lm(
  formula = dependent ~ independent + controls,
  data = my_df
)
```

## Interpreting Output {data-name="Interpreting"}

<pre>
Call:
lm(formula = military ~ industrial, data = gdp_df)

Residuals:
    Min      1Q  Median      3Q     Max 
-2.3354 -1.0997 -0.3870  0.6081  6.7508 
</pre>
<pre class="highlight-block">
Coefficients:
            Estimate Std. Error t value Pr(>|t|)  
(Intercept)  0.61969    0.59526   1.041   0.3010  
industrial   0.05253    0.02019   2.602   0.0111 *
</pre>
<pre class="highlight-below">---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
Residual standard error: 1.671 on 79 degrees of freedom
  (8 observations deleted due to missingness)
Multiple R-squared:  0.07895,   Adjusted R-squared:  0.06729 
F-statistic: 6.771 on 1 and 79 DF,  p-value: 0.01106
</pre>

## Zooming In: Coefficients

::: {.coef-table}

|     |       Estimate | Std. Error | t value | Pr(>\|t\|) | |
|-:|:-:|-|-|-|-|
| <span class="cb1">**(Intercept)**</span> | [0.61969]{.cb1 .bold} |   [0.59526]{.cb1 .bold}  | [1.041]{.cb1}  |  [0.3010]{.cb1} |  |
| <span class="cb2">**industrial**</span>  | [0.05253]{.cb2 .bold}  |  [0.02019]{.cb2 .bold} |  [2.602]{.cb2} |  [0.0111]{.cb2}  | [*]{.cb2} |
| | <span class="coef-label">$\widehat{\beta}$</span> | <span class="coef-label">Uncertainty</span> | <span class="coef-label">Test statistic</span> | <span class="coef-label">How extreme is test stat?</span> | <span class="coef-label">Statistical significance</span> |

: {tbl-colwidths="[12,12,24,24,16,12]"}

:::

$$
\widehat{y} \approx \class{cb1}{\overset{\beta_0}{\underset{\small \pm 0.595}{0.620}}} +  \class{cb2}{\overset{\beta_1}{\underset{\small \pm 0.020}{0.053}}} \cdot x
$$

## Zooming In: Significance {.crunch-title}

::: {.coef-table}

|     |       Estimate | Std. Error | t value | Pr(>\|t\|) | |
|-:|:-:|-|-|-|-|
| <span class="cb1">**(Intercept)**</span> | [0.61969]{.cb1} |   [0.59526]{.cb1}  | [1.041]{.cb1 .bold}  |  [0.3010]{.cb1 .bold} |  |
| <span class="cb2">**industrial**</span>  | [0.05253]{.cb2}  |  [0.02019]{.cb2} |  [2.602]{.cb2 .bold} |  [0.0111]{.cb2 .bold}  | [*]{.cb2 .bold} |
| | <span class="coef-label">$\widehat{\beta}$</span> | <span class="coef-label">Uncertainty</span> | <span class="coef-label">Test statistic</span> | <span class="coef-label">How extreme is test stat?</span> | <span class="coef-label">Statistical significance</span> |

: {tbl-colwidths="[12,12,24,24,16,12]"}

:::

::: columns
::: {.column width="50%"}

```{r}
#| label: t-stat-intercept
#| fig-align: center
#| fig-height: 6
library(ggplot2)
int_tstat <- 1.041
int_tstat_str <- sprintf("%.02f", int_tstat)
label_df_int <- tribble(
    ~x, ~y, ~label,
    0.25, 0.05, paste0("P(t > ",int_tstat_str,")\n= 0.3")
)
label_df_signif_int <- tribble(
    ~x, ~y, ~label,
    2.7, 0.075, "95% Signif.\nCutoff"
)
funcShaded <- function(x, lower_bound, upper_bound){
    y <- dnorm(x)
    y[x < lower_bound | x > upper_bound] <- NA
    return(y)
}
funcShadedIntercept <- function(x) funcShaded(x, int_tstat, Inf)
funcShadedSignif <- function(x) funcShaded(x, 1.96, Inf)
ggplot(data=data.frame(x=c(-3,3)), aes(x=x)) +
  stat_function(fun=dnorm, linewidth=g_linewidth) +
  geom_vline(aes(xintercept=int_tstat), linewidth=g_linewidth) +
  geom_vline(aes(xintercept = 1.96), linewidth=g_linewidth, linetype="dashed") +
  stat_function(fun = funcShadedIntercept, geom = "area", fill = cbPalette[1], alpha = 0.5) +
  stat_function(fun = funcShadedSignif, geom = "area", fill = "grey", alpha = 0.333) +
  geom_text(label_df_int, mapping = aes(x = x, y = y, label = label), size = 10) +
  geom_text(label_df_signif_int, mapping = aes(x = x, y = y, label = label), size = 8) +
  # Add single additional tick
  scale_x_continuous(breaks=c(-2, 0, int_tstat, 2), labels=c("-2","0",int_tstat_str,"2")) +
  dsan_theme("quarter") +
  labs(
    title = "t Value for Intercept",
    x = "t",
    y = "Density"
  ) +
  theme(axis.text.x = element_text(colour = c("black", "black", cbPalette[1], "black")))
```

:::
::: {.column width="50%"}

```{r}
#| label: t-stat-coef
#| fig-align: center
#| fig-height: 6
library(ggplot2)
coef_tstat <- 2.602
coef_tstat_str <- sprintf("%.02f", coef_tstat)
label_df_coef <- tribble(
    ~x, ~y, ~label,
    3.65, 0.06, paste0("P(t > ",coef_tstat_str,")\n= 0.01")
)
label_df_signif_coef <- tribble(
  ~x, ~y, ~label,
  1.05, 0.03, "95% Signif.\nCutoff"
)
funcShadedCoef <- function(x) funcShaded(x, coef_tstat, Inf)
ggplot(data=data.frame(x=c(-4,4)), aes(x=x)) +
  stat_function(fun=dnorm, linewidth=g_linewidth) +
  geom_vline(aes(xintercept=coef_tstat), linetype="solid", linewidth=g_linewidth) +
  geom_vline(aes(xintercept=1.96), linetype="dashed", linewidth=g_linewidth) +
  stat_function(fun = funcShadedCoef, geom = "area", fill = cbPalette[2], alpha = 0.5) +
  stat_function(fun = funcShadedSignif, geom = "area", fill = "grey", alpha = 0.333) +
  # Label shaded area
  geom_text(label_df_coef, mapping = aes(x = x, y = y, label = label), size = 10) +
  # Label significance cutoff
  geom_text(label_df_signif_coef, mapping = aes(x = x, y = y, label = label), size = 8) +
  coord_cartesian(clip = "off") +
  # Add single additional tick
  scale_x_continuous(breaks=c(-4, -2, 0, 2, coef_tstat, 4), labels=c("-4", "-2","0", "2", coef_tstat_str,"4")) +
  dsan_theme("quarter") +
  labs(
    title = "t Value for Coefficient",
    x = "t",
    y = "Density"
  ) +
  theme(axis.text.x = element_text(colour = c("black", "black", "black", "black", cbPalette[2], "black")))
```

:::
:::

## The Residual Plot {.crunch-title .crunch-images data-name="Diagnostics"}

::: columns
::: {.column width="50%"}

* A key **assumption** required for OLS: **"homoskedasticity"**
* Given our model
$$
y_i = \beta_0 + \beta_1x_i + \varepsilon_i
$$
the errors $\varepsilon_i$ **should not vary systematically with $i$**
* Formally: $\forall i \left[ \Var{\varepsilon_i} = \sigma^2 \right]$

:::
::: {.column width="50%"}

```{r}
#| label: residual-plot
library(broom)
gdp_resid_df <- augment(lin_model)
ggplot(gdp_resid_df, aes(x = .fitted, y = .resid)) +
    geom_point(size = g_pointsize/2) +
    geom_hline(yintercept=0, linetype="dashed") +
    dsan_theme("quarter") +
    labs(
      title = "Residual Plot for Industrial ~ Military",
      x = "Fitted Value",
      y = "Residual"
    )
```

```{r}
#| label: heteroskedastic
x <- 1:80
errors <- rnorm(length(x), 0, x^2/1000)
y <- x + errors
het_model <- lm(y ~ x)
df_het <- augment(het_model)
ggplot(df_het, aes(x = .fitted, y = .resid)) +
    geom_point(size = g_pointsize / 2) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    dsan_theme("quarter") +
    labs(
        title = "Residual Plot for Heteroskedastic Data",
        x = "Fitted Value",
        y = "Residual"
    )
```

:::
:::

## Q-Q Plot

* If $(\widehat{y} - y) \sim \mathcal{N}(0, \sigma^2)$, points would lie on 45° line:

::: columns
::: {.column width="50%"}
```{r}
#| label: qq-plot-heteroskedastic
#| fig-height: 9
ggplot(df_het, aes(sample=.resid)) +
  stat_qq(size = g_pointsize/2) + stat_qq_line(linewidth = g_linewidth) +
  dsan_theme("half") +
  labs(
    title = "Q-Q Plot for Heteroskedastic Data",
    x = "Normal Distribution Quantiles",
    y = "Observed Data Quantiles"
  )
```
:::

::: {.column width="50%"}
```{r}
#| label: qq-plot-gdp
#| fig-height: 9
ggplot(gdp_resid_df, aes(sample=.resid)) +
  stat_qq(size = g_pointsize/2) + stat_qq_line(linewidth = g_linewidth) +
  dsan_theme("half") +
  labs(
    title = "Q-Q Plot for Industrial ~ Military Residuals",
    x = "Normal Distribution Quantiles",
    y = "Observed Data Quantiles"
  )
```
:::
:::

## Multiple Linear Regression

* Notation: $x_{i,j}$ = value of independent variable $j$ for person/observation $i$
* $M$ = total number of independent variables

$$
\widehat{y}_i = \beta_0 + \beta_1x_{i,1} + \beta_2x_{i,2} + \cdots + \beta_M x_{i,M}
$$

* $\beta_j$ interpretation: a one-unit increase in $x_{i,j}$ is associated with a $\beta_j$ unit increase in $y_i$, **holding all other independent variables constant**

<!-- ## Handling Categorical Data

## Interaction Effects -->

## References

::: {#refs}
:::
