How did we measure a "good" fit in e.g. DSAN 5000? High $R^2$? Low $RSS$?

:::: {.columns}
::: {.column width="50%"}

```{r}
#| label: training-data-plot
#| fig-width: 3.666
#| fig-height: 2.5
#| fig-align: left
x <- seq(from = 0, to = 1, by = 0.1)
n <- length(x)
eps <- rnorm(n, 0, 0.04)
y <- x + eps
# But make one big outlier
midpoint <- ceiling((3/4)*n)
y[midpoint] <- 0
of_data <- tibble::tibble(x=x, y=y)
# Linear model
lin_model <- lm(y ~ x)
# But now polynomial regression
poly_model <- lm(y ~ poly(x, degree = 10, raw=TRUE))
#summary(model)
ggplot(of_data, aes(x=x, y=y)) +
  geom_point(size=g_pointsize/2) +
  labs(
    title = "Training Data",
    color = "Model"
  ) +
  theme_dsan()
```

:::
::: {.column width="50%"}

```{r}
#| label: training-linear-poly
#| fig-width: 5
#| fig-height: 2.5
#| fig-align: left
ggplot(of_data, aes(x=x, y=y)) +
  geom_point(size=g_pointsize/2) +
  geom_abline(aes(intercept=0, slope=1, color="Linear"), linewidth=1, show.legend = FALSE) +
  stat_smooth(method = "lm",
              formula = y ~ poly(x, 10, raw=TRUE),
              se = FALSE, aes(color="Polynomial")) +
  labs(
    title = "A Perfect Model?",
    color = "Model"
  ) +
  theme_dsan()
```

:::
::::

:::: {.columns}
::: {.column width="50%"}

<center>
Linear Model:
</center>

```{r,echo=TRUE}
#| label: lin-model-metrics
#| echo: true
#| code-fold: show
summary(lin_model)$r.squared
get_rss(lin_model)
```

:::
::: {.column width="50%"}

<center>
Polynomial Model:
</center>

```{r,echo=TRUE}
#| label: poly-model-metrics
#| echo: true
#| code-fold: show
summary(poly_model)$r.squared
get_rss(poly_model)
```
:::

::::

---

# Cross-Validation {data-stack-name="Cross Validation"}

## Naïve Approach: Validation Set

## Better Approach: LOOCV

## Best Approach: $K$-Fold CV

---

# Model Selection Preview {data-stack-name="Model Selection"}

## The Goal

* Maximize **generalizability** by
* Penalizing **complexity**

---

## Resampling Methods

* Cross-Validation and Bootstrap

## Model Selection

* Subset Selection
* "Complexity Penalties"
* LASSO 🤠

---
