---
title: "Week 2: Linear Regression"
subtitle: "*DSAN 5300: Statistical Learning*<br><span class='subsubtitle'>Spring 2025, Georgetown University</span>"
author: "Jeff Jacobs"
institute: "[`jj1088@georgetown.edu`](mailto:jj1088@georgetown.edu)"
bibliography: "../_DSAN5300.bib"
date: 2025-01-13
date-format: full
lecnum: 2
categories:
  - "Class Sessions"
format:
  revealjs:
    df-print: kable
    footer: "DSAN 5300-01 Week 2: {{< var w02.footer >}}"
    output-file: "slides.html"
    html-math-method: mathjax
    scrollable: true
    css: "../dsan-globals/jjstyles.css"
    slide-number: true
    echo: true
    code-fold: true
    cache: true
    include-in-header:
      text: "<link rel='stylesheet' href='https://cdn.jsdelivr.net/npm/bootstrap-icons@1.11.3/font/bootstrap-icons.min.css'><link rel='stylesheet' type='text/css' href='https://cdn.jsdelivr.net/gh/dreampulse/computer-modern-web-font@master/fonts.css'>"
    simplemenu:
      flat: true
      barhtml:
        header: "<div class='menubar'><span style='position: absolute; left: 8; padding-left: 8px;'><a href='./index.html'>&larr; Return to Notes</a></span><ul class='menu'></ul></div>"
      scale: 0.5
    revealjs-plugins:
      - simplemenu
  html:
    df-print: kable
    output-file: "index.html"
    html-math-method: mathjax
    echo: true
    code-fold: true
    cache: true
    include-in-header:
      text: "<link rel='stylesheet' href='https://cdn.jsdelivr.net/npm/bootstrap-icons@1.11.3/font/bootstrap-icons.min.css'><link rel='stylesheet' type='text/css' href='https://cdn.jsdelivr.net/gh/dreampulse/computer-modern-web-font@master/fonts.css'>"
---

::: {.content-visible unless-format="revealjs"}

<center class='mb-3'>
<a class="h2" href="./slides.html" target="_blank">Open slides in new tab &rarr;</a>
</center>

:::

# Linear Regression {data-stack-name="Linear Regression"}

* What happens to my **dependent variable** $Y$ when my **independent variable** $X$ changes by **1 unit**?
* Whenever you carry out a regression, keep the **goal** in the front of your mind:

  ::: {.callout-tip .r-fit-text title="<i class='bi bi-info-circle'></i> The Goal of Regression" icon="false"}

  **Find a line $\widehat{y} = mx + b$ that best predicts $Y$ for given values of $X$**

  :::

::: {.hidden}

```{r}
#| label: r-source-globals
source("../dsan-globals/_globals.r")
set.seed(5300)
```

:::

{{< include ../dsan-globals/_globals-tex.qmd >}}


## How Do We Define "Best"? {.smaller}

* Intuitively, two different ways to measure **how well a line fits the data**:

::: {layout="[1,1]" layout-valign="center"}

```{r}
#| label: pc-line
#| fig-width: 6
library(tidyverse)
set.seed(5321)
N <- 11
x <- seq(from = 0, to = 1, by = 1 / (N - 1))
y <- x + rnorm(N, 0, 0.2)
mean_y <- mean(y)
spread <- y - mean_y
df <- tibble(x = x, y = y, spread = spread)
ggplot(df, aes(x=x, y=y)) +
  geom_abline(slope=1, intercept=0, linetype="dashed", color=cbPalette[1], linewidth=g_linewidth*2) +
  geom_segment(xend=(x+y)/2, yend=(x+y)/2, linewidth=g_linewidth*2, color=cbPalette[2]) +
  geom_point(size=g_pointsize) +
  coord_equal() +
  xlim(0, 1) + ylim(0, 1) +
  dsan_theme("half") +
  labs(
    title = "Principal Component Line"
  )
```

```{r}
#| label: reg-line
#| fig-width: 6
ggplot(df, aes(x=x, y=y)) +
  geom_abline(slope=1, intercept=0, linetype="dashed", color=cbPalette[1], linewidth=g_linewidth*2) +
  geom_segment(xend=x, yend=x, linewidth=g_linewidth*2, color=cbPalette[2]) +
  geom_point(size=g_pointsize) +
  coord_equal() +
  xlim(0, 1) + ylim(0, 1) +
  dsan_theme("half") +
  labs(
    title = "Regression Line"
  )
```

:::

::: {.aside}
On the difference between these two lines, and why it matters, I cannot recommend @gelman_data_2007 enough!
:::

## Predictive Example: Advertising Effects {.smaller .title-12}

* **Independent variable**: $ put into advertisements
* **Dependent variable**: Sales
* **Goal**: Figure out a good way to **allocate** an advertising budget

```{r}
#| label: ad-plots
library(tidyverse)
ad_df <- read_csv("assets/Advertising.csv") |> rename(id=`...1`)
long_df <- ad_df |> pivot_longer(-c(id, sales), names_to="medium", values_to="allocation")
long_df |> ggplot(aes(x=allocation, y=sales)) +
  geom_point() +
  facet_wrap(vars(medium), scales="free_x") +
  geom_smooth(method='lm', formula="y ~ x") +
  theme_dsan() +
  labs(
    x = "Allocation ($1K)",
    y = "Sales (1K Units)"
  )
```

## Explanatory Example: Industrialization Effects {.smaller}

```{r}
#| label: mil-plot-intro
library(tidyverse)
gdp_df <- read_csv("assets/gdp_pca.csv")
mil_plot <- gdp_df |> ggplot(aes(x=industrial, y=military)) +
  geom_point(size=0.5*g_pointsize) +
  geom_smooth(method='lm', formula="y ~ x", linewidth=1) +
  theme_dsan() +
  labs(
    title="Military Exports vs. Industrialization",
    x="Industrial Production (% of GDP)",
    y="Military Exports (% of All Exports)"
  )
mil_plot
```

## Simple Linear Regression {.smaller}

* For now, we treat **Newspaper**, **Radio**, **TV** advertising separately: how much do **sales** increase per $1 into [medium]? (Later we'll consider them jointly: multiple regression)
* **Our model**:

  $$
  Y = \underbrace{\param{\beta_0}}_{\mathclap{\text{Intercept}}} + \underbrace{\param{\beta_1}}_{\mathclap{\text{Slope}}}X + \varepsilon
  $$

* This model generates **predictions** via

  $$
  \widehat{y} = \underbrace{\widehat{\beta_0}}_{\mathclap{\small\begin{array}{c}\text{Estimated} \\[-5mm] \text{intercept}\end{array}}} ~+~ \underbrace{\widehat{\beta_1}}_{\mathclap{\small\begin{array}{c}\text{Estimated} \\[-4mm] \text{slope}\end{array}}}\cdot x
  $$

* Note how these predictions will be **wrong** (unless the data is perfectly linear)
* We've accounted for this in our model (by including $\varepsilon$ term)!
* But, we'd like to find estimates $\widehat{\beta}_0$ and $\widehat{\beta}_1$ that produce the "least wrong" predictions: motivates focus on **residuals...**

## Least Squares: Minimizing Residuals {.smaller .crunch-title .crunch-quarto-figure}

What can we **optimize** to ensure these residuals are as small as possible?

:::: {.columns}
::: {.column width="60%"}

```{r}
#| label: large-resids-plot
#| fig-height: 4
N <- 21
x <- seq(from = 0, to = 1, by = 1 / (N - 1))
y <- x + rnorm(N, 0, 0.25)
mean_y <- mean(y)
spread <- y - mean_y
sim_lg_df <- tibble(x = x, y = y, spread = spread)
sim_lg_df |> ggplot(aes(x=x, y=y)) +
  geom_abline(slope=1, intercept=0, linetype="dashed", color=cbPalette[1], linewidth=g_linewidth) +
  # geom_segment(xend=x, yend=x, linewidth=g_linewidth*2, color=cbPalette[2]) +
  geom_segment(aes(xend=x, yend=x, color=ifelse(y>x,"Positive","Negative")), linewidth=1.5*g_linewidth) +
  geom_point(size=g_pointsize) +
  # coord_equal() +
  theme_dsan("half") +
  scale_color_manual("Spread", values=c("Positive"=cbPalette[3],"Negative"=cbPalette[6]), labels=c("Positive"="Positive","Negative"="Negative")) +
  labs(
    title = "Regression Line"
  )
```

:::
::: {.column width="40%"}

* Sum?

```{r}
large_sum <- sum(sim_lg_df$spread)
writeLines(fmt_decimal(large_sum))
```

* Sum of Squares?

```{r}
large_sqsum <- sum((sim_lg_df$spread)^2)
writeLines(fmt_decimal(large_sqsum))
```

:::
::::

:::: {.columns}
::: {.column width="60%"}

```{r}
#| label: small-resids-plot
#| fig-height: 4
N <- 21
x <- seq(from = 0, to = 1, by = 1 / (N - 1))
y <- x + rnorm(N, 0, 0.05)
mean_y <- mean(y)
spread <- y - mean_y
sim_sm_df <- tibble(x = x, y = y, spread = spread)
sim_sm_df |> ggplot(aes(x=x, y=y)) +
  geom_abline(slope=1, intercept=0, linetype="dashed", color=cbPalette[1], linewidth=g_linewidth) +
  # geom_segment(xend=x, yend=x, linewidth=g_linewidth*2, color=cbPalette[2]) +
  geom_segment(aes(xend=x, yend=x, color=ifelse(y>x,"Positive","Negative")), linewidth=1.5*g_linewidth) +
  geom_point(size=g_pointsize) +
  # coord_equal() +
  theme_dsan("half") +
  scale_color_manual("Spread", values=c("Positive"=cbPalette[3],"Negative"=cbPalette[6]), labels=c("Positive"="Positive","Negative"="Negative")) +
  labs(
    title = "Regression Line"
  )
```

:::
::: {.column width="40%"}

* Sum?

```{r}
small_rsum <- sum(sim_sm_df$spread)
writeLines(fmt_decimal(small_rsum))
```

* Sum of Squares?

```{r}
small_sqrsum <- sum((sim_sm_df$spread)^2)
writeLines(fmt_decimal(small_sqrsum))
```

:::
::::

## Why Not Absolute Value? {.smaller .crunch-title .crunch-code .crunch-ul .crunch-quarto-figure}

* Two feasible ways to prevent positive and negative residuals **cancelling out**:
  * **Absolute value** $\left|y - \widehat{y}\right|$ or **squaring** $\left( y - \widehat{y} \right)^2$
* But remember that we're aiming to **minimize** these residuals...
* Ghost of calculus past 😱: which is **differentiable** everywhere?

:::: {.columns}
::: {.column width="50%"}

```{r}
#| label: x2-plot
#| fig-height: 7
library(latex2exp)
x2_label <- latex2exp("$f(x) = x^2$")
ggplot(data.frame(x=c(-4,4)), aes(x=x)) +
  stat_function(fun=~ .x^2, linewidth = g_linewidth) +
  dsan_theme("quarter") +
  labs(
    title=x2_label,
    y="f(x)"
  )
```

:::
::: {.column width="50%"}

```{r}
#| label: abs-x-plot
#| fig-height: 7
# Could use facet_grid() here, but it doesn't work too nicely with stat_function() :(
ggplot(data.frame(x=c(-4,4)), aes(x=x)) +
  stat_function(fun=~ abs(.x), linewidth=g_linewidth) +
  dsan_theme("quarter") +
  labs(
    title="f(x) = |x|",
    y="f(x)"
  )
```

:::
::::

## Outliers Penalized Quadratically {.smaller}

![[Image Source](https://gregorygundersen.com/blog/2020/01/04/ols/)](images/rss.png){fig-align="center"}

## Key Features of Regression Line {.crunch-math .math-90}

* Regression line is **BLUE**: **B**est **L**inear **U**nbiased **E**stimator
* What exactly is it the "best" linear estimator of?

$$
\widehat{y} = \underbrace{\widehat{\beta}_0}_{\mathclap{\small\begin{array}{c}\text{Estimated} \\[-5mm] \text{intercept}\end{array}}} ~+~ \underbrace{\widehat{\beta}_1}_{\mathclap{\small\begin{array}{c}\text{Estimated} \\[-4mm] \text{slope}\end{array}}}\cdot x
$$

is chosen so that

::: {.smallmath}

$$
\widehat{\theta} = \left(\widehat{\beta}_0, \widehat{\beta}_1\right) = \argmin_{\beta_0, \beta_1}\left[ \sum_{x_i \in X} \left(~~\overbrace{\widehat{y}(x_i)}^{\mathclap{\small\text{Predicted }y}} - \overbrace{\expect{Y \mid X = x_i}}^{\small \text{Avg. }y\text{ when }x = x_i}\right)^{2~} \right]
$$

:::

## Where Did That $\mathbb{E}[Y \mid X = x_i]$ Come From? {.smaller .title-11}

![[Image Source](https://digfir-published.macmillanusa.com/psbe4e/psbe4e_ch10_2.html)](images/regression_normal_dist.jpg){fig-align="center"}

## But... What About All the Other Types of Vars? {.title-08}

* 5000: you saw, e.g., nominal, ordinal, cardinal vars
* 5100: you wrestled with **discrete** vs. **continuous** RVs
* Good News #1: Regression can handle **all** these types+more!
* Good News #2: Distinctions between **classification** and **regression** start to diminish as you learn fancier regression methods! (One key tool here: **link functions**)
* By end of 5300 you should have something on your toolbelt for handling most cases like *"I want to do [regression / classification], but my data is [not cardinal+continuous]"*

# Deriving the Least Squares Estimate {data-stack-name="Deriving OLS"}

## A Sketch (HW is the Full Thing) {.smaller}

* OLS for regression **without** intercept $\param{\beta_0}$: Which **line through origin** best predicts $Y$?
* (Good practice + reminder of how **restricted** linear models are!)

$$
Y = \beta_1 X + \varepsilon
$$

```{r}
#| label: random-lines-origin
library(latex2exp)
set.seed(5300)
# rand_slope <- log(runif(80, min=0, max=1))
# rand_slope[41:80] <- -rand_slope[41:80]
# rand_lines <- tibble::tibble(
#   id=1:80, slope=rand_slope, intercept=0
# )
# angles <- runif(100, -pi/2, pi/2)
angles <- seq(from=-pi/2, to=pi/2, length.out=50)
possible_lines <- tibble::tibble(
  slope=tan(angles), intercept=0
)
num_points <- 30
x_vals <- runif(num_points, 0, 1)
y0_vals <- 0.5 * x_vals + 0.25
y_noise <- rnorm(num_points, 0, 0.07)
y_vals <- y0_vals + y_noise
rand_df <- tibble::tibble(x=x_vals, y=y_vals)
title_exp <- latex2exp("Parameter Space ($\\beta_1$)")
# Main plot object
gen_lines_plot <- function(point_size=2.5) {
  lines_plot <- rand_df |> ggplot(aes(x=x, y=y)) +
    geom_point(size=point_size) +
    geom_hline(yintercept=0, linewidth=1.5) +
    geom_vline(xintercept=0, linewidth=1.5) +
    # Point at origin
    geom_point(data=data.frame(x=0, y=0), aes(x=x, y=y), size=4) +
    xlim(-1,1) +
    ylim(-1,1) +
    # coord_fixed() +
    theme_dsan_min(base_size=28)
  return(lines_plot)
}
main_lines_plot <- gen_lines_plot()
main_lines_plot +
  # Parameter space of possible lines
  geom_abline(
    data=possible_lines,
    aes(slope=slope, intercept=intercept, color='possible'),
    # linetype="dotted",
    # linewidth=0.75,
    alpha=0.25
  ) +
  # True DGP
  geom_abline(
    aes(
      slope=0.5,
      intercept=0.25,
      color='true'
    ), linewidth=1, alpha=0.8
  ) + 
  scale_color_manual(
    element_blank(),
    values=c('possible'="black", 'true'=cb_palette[2]),
    labels=c('possible'="Possible Fits", 'true'="True DGP")
  ) +
  remove_legend_title() +
  labs(
    title=title_exp
  )
```

## Evaluating with Residuals {.smaller}

:::: {.columns}
::: {.column width="50%"}

```{r}
#| label: possible-choice-1
rc1_df <- possible_lines |> slice(n() - 14)
# Predictions for this choice
rc1_pred_df <- rand_df |> mutate(
  y_pred = rc1_df$slope * x,
  resid = y - y_pred
)
rc1_label <- latex2exp(paste0("Estimate 1: $\\beta_1 \\approx ",round(rc1_df$slope, 3),"$"))
rc1_lines_plot <- gen_lines_plot(point_size=5)
rc1_lines_plot +
  geom_abline(
    data=rc1_df,
    aes(intercept=intercept, slope=slope),
    linewidth=2,
    color=cb_palette[1]
  ) +
  geom_segment(
    data=rc1_pred_df,
    aes(x=x, y=y, xend=x, yend=y_pred),
    # color=cb_palette[1]
  ) +
  xlim(0, 1) + ylim(0, 1) +
  labs(
    title = rc1_label
  )
```

:::
::: {.column width="50%"}

```{r}
#| label: choice-1-resid
gen_resid_plot <- function(pred_df) {
  rc_rss <- sum((pred_df$resid)^2)
  rc_resid_label <- latex2exp(paste0("Residuals: RSS $\\approx$ ",round(rc_rss,3)))
  rc_resid_plot <- pred_df |> ggplot(aes(x=x, y=resid)) +
    geom_point(size=5) +
    geom_hline(
      yintercept=0,
      color=cb_palette[1],
      linewidth=1.5
    ) +
    geom_segment(
      aes(xend=x, yend=0)
    ) +
    theme_dsan(base_size=28) +
    theme(axis.line.x = element_blank()) +
    labs(
      title=rc_resid_label
    )
  return(rc_resid_plot)
}
rc1_resid_plot <- gen_resid_plot(rc1_pred_df)
rc1_resid_plot
```

:::
::::

:::: {.columns}
::: {.column width="50%"}

```{r}
#| label: possible-choice-2
rc2_df <- possible_lines |> slice(n() - 9)
# Predictions for this choice
rc2_pred_df <- rand_df |> mutate(
  y_pred = rc2_df$slope * x,
  resid = y - y_pred
)
rc2_label <- latex2exp(paste0("Estimate 2: $\\beta_1 \\approx ",round(rc2_df$slope,3),"$"))
rc2_lines_plot <- gen_lines_plot(point_size=5)
rc2_lines_plot +
  geom_abline(
    data=rc2_df,
    aes(intercept=intercept, slope=slope),
    linewidth=2,
    color=cb_palette[3]
  ) +
  geom_segment(
    data=rc2_pred_df,
    aes(
      x=x, y=y, xend=x,
      yend=ifelse(y_pred <= 1, y_pred, Inf)
    )
    # color=cb_palette[1]
  ) +
  xlim(0, 1) + ylim(0, 1) +
  labs(
    title=rc2_label
  )
```

:::
::: {.column width="50%"}

```{r}
#| label: choice-2-resid
rc2_resid_plot <- gen_resid_plot(rc2_pred_df)
rc2_resid_plot
```

:::
::::

## Now the Math {.smaller}

$$
\begin{align*}
\beta_1^* = \argmin_{\beta_1}\left[ \sum_{i=1}^{n}(\widehat{y}_i - y_i)^2 \right] = \argmin_{\beta_1}\left[ \sum_{i=1}^{n}(\beta_1x_i - y_i)^2 \right]
\end{align*}
$$

We can compute this derivative to obtain:

$$
\frac{\partial}{\partial\beta_1}\left[ \sum_{i=1}^{n}(\beta_1x_i - y_i)^2 \right] = \sum_{i=1}^{n}\frac{\partial}{\partial\beta_1}(\beta_1x_i - y_i)^2 = \sum_{i=1}^{n}2(\beta_1x_i - y_i)x_i
$$

And our first-order condition means that:

$$
\sum_{i=1}^{n}2(\beta_1^*x_i - y_i)x_i = 0 \iff \beta_1^*\sum_{i=1}^{n}x_i^2 = \sum_{i=1}^{n}x_iy_i \iff \boxed{\beta_1^* = \frac{\sum_{i=1}^{n}x_iy_i}{\sum_{i=1}^{n}x_i^2}}
$$

# Doing Things With Regression {data-stack-name="Interpreting Results"}

## Regression: `R` vs. `statsmodels` {.smaller .output-75}

:::: {.columns}
::: {.column width="45%"}

<center>
**In (Base) R: `lm()`**
</center>

```{r}
#| label: lin-reg
#| echo: true
#| code-fold: show
lin_model <- lm(sales ~ TV, data=ad_df)
summary(lin_model)
```

General syntax:

``` {.r}
lm(
  dependent ~ independent + controls,
  data = my_df
)
```

:::
::: {.column width="55%"}

<center>
**In Python: `smf.ols()`**
</center>

```{python}
#| label: load-data-py
#| echo: false
import pandas as pd
ad_df = pd.read_csv("assets/Advertising.csv")
```

```{python}
#| label: lin-reg-py
#| echo: true
#| code-fold: show
import statsmodels.formula.api as smf
results = smf.ols("sales ~ TV", data=ad_df).fit()
print(results.summary(slim=True))
```

General syntax:

``` {.python}
smf.ols(
  "dependent ~ independent + controls",
  data = my_df
)
```

:::
::::

## Interpreting Output {.smaller}

:::: {.columns}
::: {.column width="50%"}

```{r}
#| label: mil-plot
#| fig-height: 8
mil_plot + theme_dsan("quarter")
```

:::
::: {.column width="50%"}

```{r}
#| label: mil-reg
gdp_model <- lm(military ~ industrial, data=gdp_df)
summary(gdp_model)
```

:::
::::

## Zooming In: Coefficients {.smaller-table .crunch-title}

::: {.coef-table}

|     |       Estimate | Std. Error | t value | Pr(>\|t\|) | |
|-:|:-:|-|-|-|-|
| <span class="cb1">**(Intercept)**</span> | [0.61969]{.cb1} |   [0.59526]{.cb1}  | [1.041]{.cb1 .bold}  |  [0.3010]{.cb1 .bold} |  |
| <span class="cb2">**industrial**</span>  | [0.05253]{.cb2}  |  [0.02019]{.cb2} |  [2.602]{.cb2 .bold} |  [0.0111]{.cb2 .bold}  | [*]{.cb2 .bold} |
| | <span class="coef-label">$\widehat{\beta}$</span> | <span class="coef-label">Uncertainty</span> | <span class="coef-label">Test stat $t$</span> | <span class="coef-label">How extreme is $t$?</span> | <span class="coef-label">Signif. Level</span> |

: {tbl-colwidths="[12,12,20,22,26,8]"}

:::

$$
\widehat{y} \approx \class{cb1}{\overset{\beta_0}{\underset{\small \pm 0.595}{0.620}}} +  \class{cb2}{\overset{\beta_1}{\underset{\small \pm 0.020}{0.053}}} \cdot x
$$

## Zooming In: Significance {.crunch-title .smaller-table .crunch-details .crunch-quarto-figure}

::: {.coef-table}

|     |       Estimate | Std. Error | t value | Pr(>\|t\|) | |
|-:|:-:|-|-|-|-|
| <span class="cb1">**(Intercept)**</span> | [0.61969]{.cb1} |   [0.59526]{.cb1}  | [1.041]{.cb1 .bold}  |  [0.3010]{.cb1 .bold} |  |
| <span class="cb2">**industrial**</span>  | [0.05253]{.cb2}  |  [0.02019]{.cb2} |  [2.602]{.cb2 .bold} |  [0.0111]{.cb2 .bold}  | [*]{.cb2 .bold} |
| | <span class="coef-label">$\widehat{\beta}$</span> | <span class="coef-label">Uncertainty</span> | <span class="coef-label">Test stat $t$</span> | <span class="coef-label">How extreme is $t$?</span> | <span class="coef-label">Signif. Level</span> |

: {tbl-colwidths="[12,12,20,22,26,8]"}

:::

::: columns
::: {.column width="50%"}

```{r}
#| label: t-stat-intercept
#| fig-align: center
#| fig-height: 6
library(ggplot2)
int_tstat <- 1.041
int_tstat_str <- sprintf("%.02f", int_tstat)
label_df_int <- tribble(
    ~x, ~y, ~label,
    0.25, 0.05, paste0("P(t > ",int_tstat_str,")\n= 0.3")
)
label_df_signif_int <- tribble(
    ~x, ~y, ~label,
    2.7, 0.075, "95% Signif.\nCutoff"
)
funcShaded <- function(x, lower_bound, upper_bound){
    y <- dnorm(x)
    y[x < lower_bound | x > upper_bound] <- NA
    return(y)
}
funcShadedIntercept <- function(x) funcShaded(x, int_tstat, Inf)
funcShadedSignif <- function(x) funcShaded(x, 1.96, Inf)
ggplot(data=data.frame(x=c(-3,3)), aes(x=x)) +
  stat_function(fun=dnorm, linewidth=g_linewidth) +
  geom_vline(aes(xintercept=int_tstat), linewidth=g_linewidth, linetype="dashed") +
  geom_vline(aes(xintercept = 1.96), linewidth=g_linewidth, linetype="solid") +
  stat_function(fun = funcShadedIntercept, geom = "area", fill = cbPalette[1], alpha = 0.5) +
  stat_function(fun = funcShadedSignif, geom = "area", fill = "grey", alpha = 0.333) +
  geom_text(label_df_int, mapping = aes(x = x, y = y, label = label), size = 10) +
  geom_text(label_df_signif_int, mapping = aes(x = x, y = y, label = label), size = 8) +
  # Add single additional tick
  scale_x_continuous(breaks=c(-2, 0, int_tstat, 2), labels=c("-2","0",int_tstat_str,"2")) +
  dsan_theme("quarter") +
  labs(
    title = "t Value for Intercept",
    x = "t",
    y = "Density"
  ) +
  theme(axis.text.x = element_text(colour = c("black", "black", cbPalette[1], "black")))
```

:::
::: {.column width="50%"}

```{r}
#| label: t-stat-coef
#| fig-align: center
#| fig-height: 6
library(ggplot2)
coef_tstat <- 2.602
coef_tstat_str <- sprintf("%.02f", coef_tstat)
label_df_coef <- tribble(
    ~x, ~y, ~label,
    3.65, 0.06, paste0("P(t > ",coef_tstat_str,")\n= 0.01")
)
label_df_signif_coef <- tribble(
  ~x, ~y, ~label,
  1.05, 0.03, "95% Signif.\nCutoff"
)
funcShadedCoef <- function(x) funcShaded(x, coef_tstat, Inf)
ggplot(data=data.frame(x=c(-4,4)), aes(x=x)) +
  stat_function(fun=dnorm, linewidth=g_linewidth) +
  geom_vline(aes(xintercept=coef_tstat), linetype="dashed", linewidth=g_linewidth) +
  geom_vline(aes(xintercept=1.96), linetype="solid", linewidth=g_linewidth) +
  stat_function(fun = funcShadedCoef, geom = "area", fill = cbPalette[2], alpha = 0.5) +
  stat_function(fun = funcShadedSignif, geom = "area", fill = "grey", alpha = 0.333) +
  # Label shaded area
  geom_text(label_df_coef, mapping = aes(x = x, y = y, label = label), size = 10) +
  # Label significance cutoff
  geom_text(label_df_signif_coef, mapping = aes(x = x, y = y, label = label), size = 8) +
  coord_cartesian(clip = "off") +
  # Add single additional tick
  scale_x_continuous(breaks=c(-4, -2, 0, 2, coef_tstat, 4), labels=c("-4", "-2","0", "2", coef_tstat_str,"4")) +
  dsan_theme("quarter") +
  labs(
    title = "t Value for Coefficient",
    x = "t",
    y = "Density"
  ) +
  theme(axis.text.x = element_text(colour = c("black", "black", "black", "black", cbPalette[2], "black")))
```

:::
:::

## The Residual Plot {.crunch-title .crunch-images data-name="Diagnostics"}

::: columns
::: {.column width="50%"}

* A key **assumption** required for OLS: **"homoskedasticity"**
* Given our model
$$
y_i = \beta_0 + \beta_1x_i + \varepsilon_i
$$
the errors $\varepsilon_i$ **should not vary systematically with $i$**
* Formally: $\forall i \left[ \Var{\varepsilon_i} = \sigma^2 \right]$

:::
::: {.column width="50%"}

```{r}
#| label: residual-plot
library(broom)
gdp_resid_df <- augment(gdp_model)
ggplot(gdp_resid_df, aes(x = industrial, y = .resid)) +
    geom_point(size = g_pointsize/2) +
    geom_hline(yintercept=0, linetype="dashed") +
    dsan_theme("quarter") +
    labs(
      title = "Residual Plot for Military ~ Industrial",
      x = "Fitted Value",
      y = "Residual"
    )
```

```{r}
#| label: heteroskedastic
x <- 1:80
errors <- rnorm(length(x), 0, x^2/1000)
y <- x + errors
het_model <- lm(y ~ x)
df_het <- augment(het_model)
ggplot(df_het, aes(x = .fitted, y = .resid)) +
    geom_point(size = g_pointsize / 2) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    dsan_theme("quarter") +
    labs(
        title = "Residual Plot for Heteroskedastic Data",
        x = "Fitted Value",
        y = "Residual"
    )
```

:::
:::

## Q-Q Plot

* If $(\widehat{y} - y) \sim \mathcal{N}(0, \sigma^2)$, points would lie on 45° line:

::: columns
::: {.column width="50%"}
```{r}
#| label: qq-plot-heteroskedastic
#| fig-height: 9
ggplot(df_het, aes(sample=.resid)) +
  stat_qq(size = g_pointsize/2) + stat_qq_line(linewidth = g_linewidth) +
  dsan_theme("half") +
  labs(
    title = "Q-Q Plot for Heteroskedastic Data",
    x = "Normal Distribution Quantiles",
    y = "Observed Data Quantiles"
  )
```
:::

::: {.column width="50%"}
```{r}
#| label: qq-plot-gdp
#| fig-height: 9
ggplot(gdp_resid_df, aes(sample=.resid)) +
  stat_qq(size = g_pointsize/2) + stat_qq_line(linewidth = g_linewidth) +
  dsan_theme("half") +
  labs(
    title = "Q-Q Plot for Industrial ~ Military Residuals",
    x = "Normal Distribution Quantiles",
    y = "Observed Data Quantiles"
  )
```
:::
:::

## Multiple Linear Regression

* Notation: $x_{i,j}$ = value of independent variable $j$ for person/observation $i$
* $M$ = total number of independent variables

$$
\widehat{y}_i = \beta_0 + \beta_1x_{i,1} + \beta_2x_{i,2} + \cdots + \beta_M x_{i,M}
$$

* $\beta_j$ interpretation: a one-unit increase in $x_{i,j}$ is associated with a $\beta_j$ unit increase in $y_i$, **holding all other independent variables constant**

## Visualizing Multiple Linear Regression

![(ISLR, Fig 3.5): *A pronounced non-linear relationship. Positive residuals (visible above the surface) tend to lie along the 45-degree line, where budgets are split evenly. Negative residuals (most not visible) tend to be away from this line, where budgets are more lopsided.*](images/3_5.svg){fig-align="center"}

## Interpreting MLR {.smaller}

:::: {.columns}
::: {.column width="50%"}

```{r}
#| label: mlr-results
#| code-fold: show
mlr_model <- lm(sales ~ TV + radio + newspaper, data=ad_df)
print(summary(mlr_model))
```

:::
::: {.column width="50%"}

* **Holding `radio` and `newspaper` spending constant**...
  * An **increase of $1K** in spending on `TV` advertising is associated with
  * An **increase in sales of about 46 units**
* **Holding `TV` and `newspaper` spending constant**...
  * An **increase of $1K** in spending on `radio` advertising is associated with
  * An **increase in sales of about 189 units**

:::
::::

## But Wait... {.smaller}

:::: {.columns}
::: {.column width="50%"}

```{r}
#| label: mlr-result-side
print(summary(mlr_model))
```

:::
::: {.column width="50%"}

```{r}
#| label: lr-result-side
lr_model <- lm(sales ~ newspaper, data=ad_df)
print(summary(lr_model))
```

:::
::::

* MLR results can be **drastically different** from SLR results, because of **correlations** (next slide)
* This is a good thing! It's how we're able to **control for** confounding variables!

## Correlations Among Features {.smaller}

```{r}
#| label: feature-corr-matrix
cor(ad_df |> select(-id))
```

* Observe how $\text{cor}(\texttt{radio}, \texttt{newspaper}) \approx 0.35$ 
* In markets where we spend more on `radio` our sales will tend to be higher...
* Corr matrix $\implies$ we spend more on `newspaper` in those same markets...
* In SLR which only examines `sales` vs. `newspaper`, we (**correctly!**) observe that higher values of `newspaper` are associated with higher values of `sales`...
* In essence, `newspaper` advertising is a surrogate for `radio` advertising $\implies$ in our SLR, `newspaper` "gets credit" for the association between `radio` and `sales`

## Another MLR Superpower: Incorporating Categorical Vars {.smaller .title-09 .math-90}

*(Preview for next week)*

:::: {.columns}
::: {.column width="50%"}

$$
Y = \beta_0 + \beta_1 \times \texttt{income}
$$

```{r}
#| label: credit-plot-slr
credit_df <- read_csv("assets/Credit.csv")
credit_plot <- credit_df |> ggplot(aes(x=Income, y=Balance)) +
  geom_point(size=0.5*g_pointsize) +
  geom_smooth(method='lm', formula="y ~ x", linewidth=1) +
  theme_dsan() +
  labs(
    title="Credit Card Balance vs. Income Level",
    x="Income ($1K)",
    y="Credit Card Balance ($)"
  )
credit_plot
```

:::
::: {.column width="50%"}

$$
\begin{align*}
Y = &\beta_0 + \beta_1 \times \texttt{income} + \beta_2 \times \texttt{Student} \\
&+ \beta_3 \times (\texttt{Student} \times \texttt{Income})
\end{align*}
$$

```{r}
#| label: credit-plot-students
student_plot <- credit_df |> ggplot(aes(x=Income, y=Balance, color=Student)) +
  geom_point(size=0.5*g_pointsize) +
  geom_smooth(method='lm', formula="y ~ x", linewidth=1) +
  theme_dsan() +
  labs(
    title="Credit Card Balance vs. Income Level",
    x="Income ($1K)",
    y="Credit Card Balance ($)"
  )
student_plot
```

:::
::::

* Why do we need the $\texttt{Student} \times \texttt{Income}$ term?
* Understanding this setup will open up a **vast** array of possibilities for regression 😎

## References

::: {#refs}
:::
